Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2023-04-20 main.tex
worked on results section
first attempt at conclusion
minor edits

\title{Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation}
\author{Timo Betcke\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, Micha\l{} Bosy\thanks{School of Computer Science and Mathematics, Kingston University London, Penrhyn Road, Kingston upon Thames, KT1 2EE, UK}, Erik Burman\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, Christopher D. Cooper\thanks{Department of Mechanical Engineering, Universidad T\'ecnica Federico Santa Mar\'ia, Valpara\'iso 2390123, Chile}, Matthew W. Scroggs\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}}

@@

\section*{\sffamily \Large RESULTS}
\input{results}
%((Place Results here. Not needed for review articles.))

\section*{\sffamily \Large First-order heading}

@@

\subsection*{\sffamily \large Second-order heading}

\subsubsection*{\sffamily \normalsize Third-order heading}

{\sffamily \small Fourth-order heading}\\

\section*{\sffamily \Large DISCUSSION}

@@

\section*{\sffamily \Large CONCLUSIONS}

%((Place Conclusions here.))

>>>>

\title{Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation}
\author{Timo Betcke\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, Micha\l{} Bosy\thanks{School of Computer Science and Mathematics, Kingston University London, Penrhyn Road, Kingston upon Thames, KT1 2EE, UK}, Erik Burman\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, Christopher D. Cooper\thanks{Department of Mechanical Engineering and Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso, Universidad T\'ecnica Federico Santa Mar\'ia, Valpara\'iso 2390123, Chile}, Matthew W. Scroggs\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}}

@@

\section*{\sffamily \Large RESULTS AND DISCUSSION}
\input{results}
%((Place Results here. Not needed for review articles.))

%\section*{\sffamily \Large First-order heading}

@@

%\subsection*{\sffamily \large Second-order heading}

%\subsubsection*{\sffamily \normalsize Third-order heading}

%{\sffamily \small Fourth-order heading}\\

%\section*{\sffamily \Large DISCUSSION}

@@

\section*{\sffamily \Large CONCLUSIONS}
\input{conclusion}
%((Place Conclusions here.))

### 2023-04-20 results.tex
minor edits

For very large grid sizes the user can enable FMM assembly which internally is handled in Bempp-cl through an interface to the Exafmm-t FMM library. For $N$ surface elements this reduces the memory and computational complexity from $\mathcal{O}(N^2)$ in the dense assembly case to $\mathcal{O}(N)$ in the FMM case, making large boundary element problems tractable on standard workstations.

@@

The solvation energy computed with the two schemes is presented by Figure \ref{fig:arg_constant_energy}, which, as expected, converges to a similar answer as the mesh is refined.
Then, Figure \ref{fig:arg2_constant_time_iter} compares the iteration count and time-to-solution. The left plot shows that using only BEM outperforms the coupled FEM-BEM approach in terms of iterations count. However, if we look at the total time that solvers take to obtain the solution, we can see the advantage of using the FEM-BEM coupling. The higher computational cost is caused by the need of using a hypersingular operator in the BEM formulation. \textcolor{red}{Chris: I thought BEM-BEM outperformed FEM-BEM in total time. Am I wrong there?} \textcolor{blue}{Michal: Not for this formulation of BEM-BEM since we have hypersingular operator. But you can see that for denser meshes we should outperform it.}

@@

\begin{figure}
\centering
   \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_iter.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_time.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_setup_time.png}
  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_total_time.png}
%   \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_iter_11.png}
%%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_time_11.png}
%%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_setup_time_11.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_total_time_11.png}
  \caption{Iteration count (left) and time-to-solution (right) for Arginine with a constant permittivity. %(left: Online time taken to solve systems, right: Offline time taken to set up systems). 
%NOTE: x axis differ between top and bottom. Also, should we include preconditioned vs non preconditioned? We could easily precondition BEM-BEM too.  %(1) fix title, (2) can we put them in a single plot, (3) 
%maybe we could also add a "error" wrt extrapolation, or a plot with energy for each mesh?
}
\label{fig:arg2_constant_time_iter}
\end{figure}

@@

So far, we have only tested the FEM-BEM coupled approach with small structures. Here, we study its behaviour with larger structures to evaluate its applicability in more realistic settings, and  in test cases where BEM-BEM coupling is not an alternative. Using the same Gaussian permittivity field inside the molecule, we computed the solvation free energy of protein G B1 (955 atoms, PDB code 1pgb), lysozyme (1960 atoms, PDB code 1lyz), the barnase-barstar complex (9464 atoms, PDB code 1x1u), and immunoglobulin G (20148 atoms, PDB code 1igt). All structures were parameterized with the Amber\cite{Swanson05} force field using pdb2pqr.\cite{Dolinsky04} The surface was meshed with Nanoshaper\cite{decherchi2013general} considering a grid scale of 1.5, and then used to mesh the solute region with pyGAMer\cite{lee2020open} and TetGen\cite{hang2015tetgen} with a radius-edge ratio of 1.0. Solvation free energy and timings for these runs are presented in Table \ref{table:large_variable}, showing that our FEM-BEM coupling method can reach medium-to-large sized proteins on a workstation.

Table \ref{table:large_variable} shows that the iteration count increases with the problem size. This is expected for Johnson-N\'ed\'elec, which is the simplest coupling strategy. To isolate the effect of the increased iterations in the analysis, we separate timings in ``setup'' and ``solving'' time, where the ``setup'' time is independent of the number of iterations, and the ``solving'' time corresponds to the time spent in the GMRES solver. Then, the time-per-iteration in table \ref{table:large_variable} only considers the ``solving'' time. Looking at the increase in degrees of freedom, which scales equivalently for FEM and BEM, the ``setup'' time scales slightly worse than $\mathcal{O}(N^2)$, mainly due to the dense nature of the BEM portion of the matrix \textcolor{red}{(check if not using fmm  \textcolor{blue}{we are using FMM})}. On the other hand, the ``solving time per iteration'' scales closer to $\mathcal{O}(N)$. This is an indication that the high ``solving time'' is mainly due to the increase in iteration count, and having better conditioned coupling methods, such as the so-called hybrid approach,\cite{betcke2022hybrid} can have a large impact in the time to solution.

Even though this scheme is capable of calculating the electrostatic potential in medium-to-large proteins, the largest test case in Table \ref{table:large_variable} (1igt) used up most of the available RAM memory (\textcolor{red}{check and rewrite \textcolor{blue}{we are using around 50\% of memory}}). If we were aiming at larger structures, such as full viruses,\cite{MartinezETal2019,wang2021high} we would require to improve not only in the conditioning of the system, but also in the choice of fast algorithms\cite{wang2021exafmm,kailasa2023pyexafmm} and parallelization.

>>>>

For very large grid sizes the user can enable fast multipole method (FMM) assembly which internally is handled in Bempp-cl through an interface to the Exafmm-t FMM library. For $N$ surface elements this reduces the memory and computational complexity from $\mathcal{O}(N^2)$ in the dense assembly case to $\mathcal{O}(N)$ in the FMM case, making large boundary element problems tractable on standard workstations.

@@

The solvation energy computed with the two schemes is presented by Figure \ref{fig:arg_constant_energy}, which, as expected, converges to a similar answer as the mesh is refined.
Figure \ref{fig:arg2_constant_time_iter} compares the iteration count and time-to-solution. The left plot shows that using only BEM outperforms the coupled FEM-BEM approach in terms of iterations count. However, if we look at the total time that solvers take to obtain the solution, we can see the advantage of using the FEM-BEM coupling. The higher computational cost is caused by the need of using a hypersingular operator in the BEM formulation, and the fact that we are not using any acceleration method ($ie.$ FMM \textcolor{red}{true?}). Regardless, timings for FEM-BEM coupling scale much worse with size than its pure BEM counterpart, indicating that BEM-BEM coupling is better for larger problems. 
%\textcolor{red}{Chris: I thought BEM-BEM outperformed FEM-BEM in total time. Am I wrong there?} \textcolor{blue}{Michal: Not for this formulation of BEM-BEM since we have hypersingular operator. But you can see that for denser meshes we should outperform it.}

@@

\begin{figure}
\centering
   \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_iter.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_time.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_setup_time.png}
  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine2_const_coeff_total_time.png}
%   \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_iter_11.png}
%%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_time_11.png}
%%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_setup_time_11.png}
%  \includegraphics[width=0.45\linewidth]{DolfinX_Arginine_const_coeff_total_time_11.png}
  \caption{Iteration count (left) and time-to-solution (right) for arginine with a constant permittivity. %(left: Online time taken to solve systems, right: Offline time taken to set up systems). 
%NOTE: x axis differ between top and bottom. Also, should we include preconditioned vs non preconditioned? We could easily precondition BEM-BEM too.  %(1) fix title, (2) can we put them in a single plot, (3) 
%maybe we could also add a "error" wrt extrapolation, or a plot with energy for each mesh?
}
\label{fig:arg2_constant_time_iter}
\end{figure}

@@

So far, we have only tested the FEM-BEM coupled approach with small structures. Here, we study its behaviour with larger structures to evaluate its applicability in more realistic settings, and  in test cases where BEM-BEM coupling is not an alternative. Using the same Gaussian permittivity field inside the molecule, we computed the solvation free energy of protein G B1 (955 atoms, PDB code 1pgb), lysozyme (1960 atoms, PDB code 1lyz), the barnase-barstar complex (9464 atoms, PDB code 1x1u), and immunoglobulin G (20148 atoms, PDB code 1igt). All structures were parameterized with the Amber\cite{Swanson05} force field using pdb2pqr.\cite{Dolinsky04} The surface was meshed with Nanoshaper\cite{decherchi2013general} considering a grid scale of 1.5, and then used to mesh the solute region with pyGAMer\cite{lee2020open} and TetGen\cite{hang2015tetgen} with a radius-edge ratio of 1.0. Solvation free energy and timings for these runs are presented in Table \ref{table:large_variable}, showing that our FEM-BEM coupling method can reach medium-to-large sized proteins on a workstation. For these larger structures we enabled the FMM capabilities of Bempp-cl.

Table \ref{table:large_variable} shows that the iteration count increases with the problem size. This is expected for Johnson-N\'ed\'elec, which is the simplest coupling strategy. To isolate the effect of the increased iterations in the analysis, we separate timings in ``setup'' and ``solving'' time, where the ``setup'' time is independent of the number of iterations, and the ``solving'' time corresponds to the time spent in the GMRES solver. Then, the time-per-iteration in Table \ref{table:large_variable} only considers the ``solving'' time. Looking at the increase in degrees of freedom (DOFs), which grows equivalently for FEM and BEM, the ``setup'' time scales slightly worse than $\mathcal{O}(N^2)$, but represents the smalle fraction of the total time. On the other hand, the ``solving time per iteration'' scales closer to $\mathcal{O}(N)$, which is expected, as we are using a FMM for the BEM portion of the matrix. This is an indication that the high ``solving time'' is mainly due to the increase in iteration count, and having better conditioned coupling methods, such as the so-called hybrid approach,\cite{betcke2022hybrid} can have a large impact in the time to solution.

Even though this scheme is capable of calculating the electrostatic potential in medium-to-large proteins, the largest test case in Table \ref{table:large_variable} (1igt) used up half of the available RAM memory \textcolor{red}{check}. 
%(\textcolor{red}{check and rewrite \textcolor{blue}{we are using around 50\% of memory}}). 
If we were aiming at larger structures, such as full viruses,\cite{MartinezETal2019,wang2021high} we would require to improve not only in the conditioning of the system, but also in the optimization of our fast algorithms\cite{wang2021exafmm,kailasa2023pyexafmm} and parallelization.
